---
title: "Cluster.identification"
author: "Jan Lanzer"
date: "7/27/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
library(tidyverse)
#library(progeny)
library(dorothea)
library(decoupleR)
library(cowplot)
library(pheatmap)

data.all= readRDS("data/data.limma.rds")

data.filt= readRDS("data/data.limma_filt.rds")
```

# patient distance and filtering of outliers
```{r}
library(lsa)

#use cosine similarity on all GEX, all samples:
voom_count= data.all$GEX$voom_count
ds. = cosine(voom_count)
pheatmap(ds.)

#plot clustering on all GEX, outliers removed:
voom_count= data.filt$GEX$voom_count
ds. = cosine(voom_count)
pheatmap(ds.)

# #eucledian:
# ds. = dist(t(voom_count))
# pheatmap(ds.)
# ds.


```

# Find cluster based on gene sets 
1 use prior knowldege gene set 
2 use NABA (fibrosis)
```{r}

#load gene set with AF prior knowledge: 
genes_felix = read.csv("data/genes_felix.csv")[,1]
prior.k= enframe(genes_felix) %>%
  mutate(mor = 1, likelihood= 1, gene.set= "prior")

  
#load  NABA
processNABA = function(filepath = "/home/jan/R-projects/sc-exploration/data/NABAgsets.xls") {
  con = file(filepath, "r")
  naba_gsets = list()
  while ( TRUE ) {
    line = readLines(con, n = 1)
    if ( length(line) == 0 ) {
      break
    }
    split_line = unlist(strsplit(line,split="\t"))
    naba_gsets[[split_line[1]]] = split_line[3:length(split_line)]
  }
  close(con)
  return(naba_gsets)
}

naba_gsets= processNABA()

x = map(names(naba_gsets) , function(y){
  enframe(naba_gsets[[y]]) %>% mutate(gene.set= y)
  })

naba_gsets2= do.call(rbind, x) %>%
  mutate(mor = 1, likelihood= 1)

all_genesets= rbind(naba_gsets2, prior.k)

#enrich_ 
.R= decouple(data.filt$GEX$voom_count, network = all_genesets, .source = gene.set, .target = value, statistics = c("gsva")) 

#plot contrast AF-CT 
.R %>% 
  left_join(data.all$target %>% rename(condition= sampleID)) %>% 
  ggplot(.,aes(x=tf, y= score, fill = group))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle= 60, hjust= 1
  ))

# Identifying AF clusters with NABA scores
AF.NABA= .R %>% 
  left_join(data.all$target %>% rename(condition= sampleID)) %>%
  filter(group == "AF") %>% 
  select(condition, score, tf) %>%
  pivot_wider( names_from= condition, values_from = score) %>% column_to_rownames("tf")

pheatmap(AF.NABA)

```

samples 1.28, 1-29, 1.26 and 1.27 show high fibrosis and low gene expression of the submitted prior

# now we will compare actual gene expression values for the genes of interest

```{r}

AF.patients= data.filt$target %>% filter(group== "AF") %>% pull(sampleID)

x= voom_count[grepl("KCNK", rownames(voom_count)),AF.patients]
pheatmap(x)

sort(voom_count["KCNK3",AF.patients])
```


# now  we will cluster AF patients based on TF activity
```{r message=FALSE, warning=FALSE}
data= data.filt

target= data$target
data(dorothea_hs, package = "dorothea")

regulons <- dorothea_hs %>%
  dplyr::filter(confidence %in% c("A", "B","C"))

# some ETV TFs are low confidence. we add them manually to the selection
dorothea_hs %>% filter(grepl("ETV", tf)) %>% distinct(tf, confidence)

regulons.extended = rbind(regulons, dorothea_hs %>% filter(grepl("ETV", tf), 
                                                           !confidence %in% c("A", "B","C")) 
) %>% mutate(label= paste0(tf,"_", confidence)) %>%
  select(-tf)%>% rename(tf= label)
#scale(data$GEX$voom_count,center = T, scale = T)

regulons.extended = rbind(regulons, dorothea_hs %>% filter(grepl("ETV", tf), 
                                                           !confidence %in% c("A", "B","C")) 
) %>% mutate(likelihood=1)

##choose mehtod

meth= c("scira", "gsva")

dec.res= decouple(scale(data$GEX$voom_count,center = T, scale = T), network = regulons.extended,statistics = meth, .source =tf, .target = target )

# dec.res= decouple(data$GEX$voom_count, network = regulons.extended,statistics =  c("gsva", "mean", "pscira", "scira", "viper"), .source =tf, .target = target )

map(c("gsva", "scira")[1], function(x){
dec.res.m = dec.res %>%
  filter(statistic== x, 
         grepl("AF", condition))%>%
  pivot_wider(id_cols= c(condition, score, tf), names_from = condition, values_from  = score) %>% 
  as.data.frame() %>%
 column_to_rownames("tf")%>% as.matrix

pheatmap(dec.res.m)
pheatmap(dec.res.m[grepl("ETV", rownames(dec.res.m)), ])

})


```

 Patients  1.28, 1-29, 1.26 and 1.27  show high ETV4, ETV6, and ETV7 activity. Same patients as in fibrosis analysis
 Careful, some ETV regulons are inferred and might be inacurate. Patient 19_AF is a bit inbetween 

# Next we will use these tow patient groups of AF to test for clinical variables.
```{r}
cluster.fib= c("1.28_AF", "1.29_AF", "1.26_AF","1.27_AF")
target= data.filt$target
target = target %>% filter(group =="AF") %>% mutate(cluster= ifelse(sampleID %in% cluster.fib, "fibrotic", "non.fibrotic"))

#which(colnames(target) == "AFstage")
fact.target= target[,37:length(colnames(target))] %>% mutate_all(as.factor) %>% select(-cluster)

test.res= map(colnames(fact.target), function(x){
  #print(x)
  if(length(unique(target[[x]]))>1){
    test.res= fisher.test(target[[x]], target$cluster)
    get.it= c(test.res$p.value, test.res$estimate)
    return(get.it)
  }
  return(NULL)
})

names(test.res) = colnames(fact.target)
test.res= do.call(rbind, test.res) 
colnames(test.res)= c("p.val", "odds.ratio")
test.res= test.res%>%as.data.frame()  %>% rownames_to_column("feature") %>% as_tibble()  %>%  mutate(logp= -log10(p.val))
test.res %>% arrange(desc(logp)) %>% print(n=200)
#test.res= fisher.test(x= target$cluster, y= target$CABG)



### now run ANOVas


cont.target= target[,1:35] %>% select(-sampleID)

ttest.res= map(colnames(cont.target), function(x){
  #print(x)
  if( is.numeric(target[[x]])){
    fib.samples= target %>% filter(sampleID %in% cluster.fib)
    n.fib=  target %>% filter(!sampleID %in% cluster.fib)
    if(sum(!is.na(fib.samples[[x]]))>2 & sum(!is.na(n.fib[[x]]))>2){
      test.res= t.test( n.fib[[x]], fib.samples[[x]])
      return(c(test.res$statistic, test.res$p.value))
      #get.it= c(test.res$p.value, test.res$estimate)
      
    }
    #return(get.it)
  }
  return(NULL)
})
names(ttest.res) = colnames(cont.target)
ttest.res= do.call(rbind, ttest.res) 
colnames(ttest.res)= c("t", "p.val")
ttest.res= ttest.res%>%as.data.frame()  %>% rownames_to_column("feature") %>% as_tibble()  %>%  mutate(logp= -log10(p.val))
ttest.res %>% arrange(desc(logp)) %>% print(n=200)
#test.res= fisher.test(x= target$cluster, y= target$CABG)

#print(target %>% select(group, cluster, everything()))
```

